---
title: 'Final Project Proposal: COVID-19 Dataset'
author: 'The Lads: Frankie Willard, Manny Mokel, Alex Katopodis, Parker Dingman'
subtitle: Due Friday, October 9, 11:59 PM
output:
  html_document: default
  pdf_document: default
---

### Section 1- Introduction

Throughout the year 2020, the COVID-19 pandemic took the world by storm, deeply impacting every country on the planet, albeit with differing degrees of severity. As cases continued to rise, families suffered from the loss of family members, jobs, social interactions, disposable income, and more. This public health crisis became severe enough such that many countries took decisive action, shutting down their economies to prioritize the lives of the citizens of their country. Meanwhile, other countries were less strict in their policies, attempting to preserve their economy at the potential expense of their citizen’s lives. The difference in each country’s characteristics, demographics, public health capacities, and the strictness of COVID-19 policies led to vastly different effects of the pandemic on different countries. Given our personal connections to the effects of the pandemic through our lives, our friends, and our families, we wanted to determine what led to the pandemic affecting some places worse than others.  
For our final project, we will be investigating country-level COVID-19 data to determine the relationship between their characteristics and demographics to virus transmission and deaths. By analyzing the effects of different country characteristics, we seek to determine specifically which variables are associated with stringency indexes, cases per capita, and deaths per case.

Research Question:

How do a country's characteristics, geography, and  demographics impact the strictness of their COVID policy, as well as the total spread and effects of COVID-19?

Hypotheses:

We hypothesize that cases per capita will have a strong negative correlation to stringency index.

We hypothesize that deaths per case will be largely determined by the GDP per capita, the number of citizens aged 65+, hospital beds per thousand, the human development index, and columns concerning pre-existing conditions (i.e. diabetes prevalence, cardiovascular death rate, etc).

We hypothesize that the stringency index is going to vary the most by continent. We expect Europe, Asia, Oceania, and North America to have higher stringency indexes, while Africa and South America will have lower stringency indexes. 

### Section 2- Data description

We selected a data set from Our World in Data. Each observation in the data set shows relevant COVID-19 data for a particular country on a given date. The COVID-19 data in the data set includes total deaths, total cases, new deaths, new cases, total cases per million, total deaths per million, total tests, new tests, total tests per thousand, positive rate, as well as telling country numbers such as stringency index (composite measure of government strictness policy) and hospital beds per thousand. Additionally, the data set includes country characteristics including population density, median age, GDP per capita, diabetes prevalence, life expectancy, and extreme poverty rate. While the previous variables are quantitative, the data set also includes categorical variables when it comes to geography such as the country and continent. 

How The Data Was Originally Collected:
“Our World In Data” uses data from the European Center for Disease Prevention and Control (ECDC), a world leader for COVID-19 data. The ECDC has a team of epidemiologists that works every day to screen up to 500 sources to get the latest figures. These sources include ministries of health (43%),  websites of public health institutes (9%),  websites of public health institutes (6%), World Health Organization (WHO) websites,  WHO situation reports (2%), and official dashboards and interactive maps from national and international institutions (10%). The EDEC also utilizes social media accounts maintained by national authorities, ministries of health, and official media outlets (30%). These social media sources are screened and validated by the other sources mentioned previously. The data is recorded daily, and we will be using the dataset updated as of October 9, 2020 (10:30, London time).

Sources:
https://ourworldindata.org/coronavirus-source-data
https://www.ecdc.europa.eu/en/covid-19/data-collection


### Section 3- Glimpse of data

```{r glimpse-data, warning=FALSE, message=FALSE}
library(tidyverse)

covid <- read_csv("data/covid-data.csv")
glimpse(covid)
```

```{r Add GG Inference}

install.packages("gginference")

library(gginference)

```



### Draft Report



###Exploratory Data Analysis

```{r Scatterplot}

covid_final_date <- covid %>% 
  filter(date == "2020-10-05" & !is.na(continent)) %>% 
  print(covid_final_date)


ggplot(data = covid_final_date, mapping = aes(x = continent , 
                                              y = total_cases)) + 
  geom_boxplot()


```


``` {r}

covid_stringency <- covid_final_date %>%
  filter(!is.na(stringency_index)) %>%
  arrange(desc(stringency_index)) %>%
  select(stringency_index, location) %>%
  print(covid_stringency)

# US Stringency Index is 66

```



### TO-DO
1. Create categorical variables (GDP, Stringency, Etc.)
2. Figure out which variables have the most impact on Stringency index, cases per capita, cases per death, etc.

Manny and Frankie: H_2

Alex and Parker: H_1 and H_3

H_1: We hypothesize that cases per capita will have a strong negative correlation to stringency index.

H_2: We hypothesize that deaths per case will be largely determined by the GDP per capita, the number of citizens aged 65+, hospital beds per thousand, the human development index, and columns concerning pre-existing conditions (i.e. diabetes prevalence, cardiovascular death rate, etc).

H_3: We hypothesize that the stringency index is going to vary the most by continent. We expect Europe, Asia, Oceania, and North America to have higher stringency indexes, while Africa and South America will have lower stringency indexes.


Hypothesis 1:


``` {r general-correlations}

covid <- covid %>%
  mutate(new_cases_per_cap = (new_cases/population),
         new_deaths_per_cap = (new_deaths/population),
         total_cases_per_cap = (total_cases/population),
         total_deaths_per_cap = (total_deaths/population))

covid_num <- select_if(covid, is.numeric) # Gets all numeric variables
covid_cat <- select_if(covid, is.character) # Gets all categorical variables
covid_cat$stringency_index = covid$stringency_index

col_names <- colnames(covid_num)
cors <- c(rep(NA, ncol(covid_num)))

for(i in 1:ncol(covid_num)) {
  isolated <- covid_num %>%
    filter(!is.na(stringency_index) & !is.na(covid_num[[i]])) %>%
    select(stringency_index, col_names[i])
  
  cors[i] <- cor(isolated)[2]
}

correlations <- tibble(cors)
correlations$other_var = col_names
correlations <- correlations %>%
  filter(other_var != "stringency_index")%>% 
  arrange(desc(abs(cors)))


correlations

```

$H_0$: The true correlation of a country's stringency index and total cases per capita is zero. There is no correlation. ($\rho$ = 0)

$H_1$: The true correlation of a country's stringency index and total cases per capita is not zero. There is a negative correlation. ($\rho <$ 0)

$\alpha$: 0.05

```{r Bootstrap for Correlations}

covid %>% 
  filter(!is.na(stringency_index) & !is.na(total_cases_per_cap)) %>% 
  summarize(cor_stringency = cor(stringency_index, total_cases_per_cap))

cor.test(covid$stringency_index, covid$total_cases_per_cap, alternative="less")

```

In summarizing the correlation between stringency index and total cases per capita, it becomes clear that our hypothesis is likely incorrect.

In performing a correlation test using Central Limit Theorem, we have a t distribution with test statistic t = 29.683 and 40315 degrees of freedom. Because our calculated p-value of 1 is greater than my predetermined $\alpha$ of 0.05, we fail to reject the null hypothesis. There is not enough evidence to suggest any negative correlation between stringency index and total cases per capita.



```{r Visualize Total Cases per Capita}

covid_continental_stringency <- covid %>%
  group_by(date) %>%
  filter(!is.na(stringency_index) & !is.na(total_cases_per_cap)) %>%
  mutate(avg_str_index = mean(stringency_index),
         avg_total_cases_cap = mean(total_cases_per_cap),
         avg_new_cases_cap = mean(new_cases_per_cap))

#Make title more informative
ggplot(data = covid_continental_stringency, aes(x=date, y = avg_str_index)) +
  geom_line()+
  labs(x = "Date",
       y = "Mean Stringency Index",
       title = "Mean Stringency Index Over Time")+
ggplot(data = covid_continental_stringency, aes(x=date, y= avg_total_cases_cap)) +
  geom_line()+
  labs(x = "Date",
       y = "Mean Total Cases Per Capita",
       title = "Mean Total Cases Per Capita Over Time")  

```

In visualizing a graph of the average global stringency index over time alongside a graph of the average total cases per capita of countries over time, it becomes clear that our hypothesis seems ill-advised. While governments initially increased their stringency indexes greatly, there reaches a point in mid-April where it begins decreasing, while total cases per capita retains exponential growth until October.

-------------------Insert explanation for why this may be---------------

In calculating our summary statistic for correlation between stringency index and total cases per capita, we see that they appear to have a positive correlation. This seems to be supported by the graph. The reason for this may be because countries started with extremely low stringency indexes and were reactive in their COVID responses, such that COVID infiltrated their countries before their stringency index became so high. Thus, they would have higher stringency indexes and higher case counts, as they failed to prevent the spread and increased their stringency index as a result of cases becoming so high. Thus, total cases per capita increased at a high rate despite stringency index being high, and thus it appears to have a positive correlation.


Thus, we would like to investigate whether there may be a statistically significant positive correlation between total cases per capita and stringency index, as well as new cases per capita and stringency index.  

```{r Visualize Stringency Index with New Cases Per Capita}

#Make title more informative
ggplot(data = covid_continental_stringency, aes(x=date, y = avg_str_index)) +
  geom_line()+
  labs(x = "Date",
       y = "Mean Stringency Index",
       title = "Mean Stringency Index Over Time")+
ggplot(data = covid_continental_stringency, aes(x=date, y= avg_new_cases_cap)) +
  geom_line()+
  labs(x = "Date",
       y = "Mean New Cases Per Capita",
       title = "Mean New Cases Per Capita Over Time")  


```


```{r Summarize Correlation Stringency New Cases Per Capita Stringency}

covid %>% 
  filter(!is.na(stringency_index) & !is.na(new_cases_per_cap)) %>% 
  summarize(cor_stringency_new_cases = cor(stringency_index, new_cases_per_cap))

# Exploration of Hypothesis Test 1
cor.test(covid$stringency_index, covid$new_cases_per_cap, alternative="less")

```



#dataset$rate <- 0
#for(i in 2:nrow(dataset)){
#dataset$rate[i] <- dataset$cases[i]/dataset$cases[i - 1]
#}

```{r Cor Test}


covid_by_country <- covid %>%
  group_by(location) %>%
  filter(!is.na(total_cases)) %>%
  filter(!is.na(new_cases)) %>%
  filter(!is.na(stringency_index))

covid_by_country$growth_total_cases <- 0
covid_by_country$growth_new_cases <- 0

# This Loop Takes a While but it WILL run
for (i in 2:nrow(covid_by_country)) {
  covid_by_country$growth_total_cases[i] <- covid_by_country$total_cases[i] / covid_by_country$total_cases[i - 1]
  covid_by_country$growth_new_cases[i] <- covid_by_country$new_cases[i] / covid_by_country$new_cases[i - 1]
}

```

``` {r}
covid_by_country %>%
  filter(date > "2020-04-15")

covid_by_country %>%
  filter(date > "2020-04-15" & 
           is.finite(growth_total_cases) &
           is.finite(growth_new_cases)) %>%
  summarize(cor_tot = cor(growth_total_cases, stringency_index),
            cor_new = cor(growth_new_cases, stringency_index))

```

Hypothesis 3




$H_0$: The true mean stringency index of Europe, Asia, Oceania, and North America is equal to the true mean stringency index of Africa and South America

$H_1$: The true mean stringency index of Europe, Asia, Oceania, and North America is greater the true mean stringency index of Africa and South America

We are performing a CLT-based Hypothesis Test at the $\alpha$ = .05 level.

``` {r parker}
# We hypothesize that the stringency index is going to vary the most by
# continent. We expect Europe, Asia, Oceania, and North America to have higher
# stringency indexes, while Africa and South America will have lower stringency
# indexes.

#Change from covid to covid final date- shouldnt be using all observations
#Could be interesting to note how countries stringency indexes change




# covid_continent_str_i_2 <- covid %>%
#   group_by(continent) %>%
#   filter(!is.na(stringency_index)) %>%
#   mutate(avg_str_index = mean(stringency_index)) %>%
#   select(continent, avg_str_index) %>%
#   summarise_all(mean) %>%
#   arrange(desc(avg_str_index))
#   print(covid_continent_str_i_2)


covid_si <- covid_final_date %>%
  filter(!is.na(stringency_index)) %>%
  mutate(grouped_continents = case_when(
    ((continent == "Africa") | (continent == "South America")) ~ "af_sa",
    ((continent == "Europe") | (continent == "Asia") ||
       (continent == "Oceania") | (continent == "North America"))
    ~ "eu_as_oc_na"))
 
#Group 1
covid_af_sa <- covid_si%>% 
  filter(grouped_continents == "af_sa") %>% 
  summarise(mean_si = mean(stringency_index)) %>% 
  print(covid_af_sa)

#Group 2
covid_eu_as_oc_na <- covid_si%>% 
  filter(grouped_continents == "eu_as_oc_na") %>% 
  summarise(mean_si = mean(stringency_index)) %>% 
  print(covid_eu_as_oc_na)

t.test(stringency_index ~ grouped_continents,
       data = covid_si,
       mu = 0,
       var.equal = FALSE,
       alternative = "less",
       conf.level = 0.95)

  
  
  
  # isolated <- covid_num %>%
  #   filter(!is.na(deaths_per_case) & !is.na(covid_num[[i]])) %>%
  #   select(col_names[i], deaths_per_case)
  # 
  # cors_deaths[i] <- cor(isolated)[2]

  
  
```

Using Central Limit Theorem, I have a t distribution with test statistic t = 34.081 and 29017 degrees of freedom.

We have a P-value = 1 > $\alpha$ = 0.05. Thus, we fail to reject $H_0$. We do not have sufficient evidence to suggest that the true mean stringency index of Europe, Asia, Oceania, and North America is greater the true mean stringency index of Africa and South America.

In looking at the summary statistics for the dataset, it becomes clear our perception of global stringency indexes by continent was misguided, as South America had the highest mean stringency index and Africa had the 4th highest. Thus, the average stringency index of Africa and South America was 55.231, which is greater than the mere 50.87269 of Europe, Asia, Oceania, and North America.
 


```{r Further Exploration of Hypothesis 3}

chisq.test(table(covid$continent, covid$stringency_index))

```

While we may have determined that we did not have sufficient evidence to suggest that true mean stringency index of Europe, Asia, Oceania, and North America is greater the true mean stringency index of Africa and South America, we see from the summary statistics that different continents do have different mean stringency indexes. Thus, we would like to determine if continent has a statistically significant effect on stringency index.

$H_0$: Stringency index and continent are independent (have no association).
$H_1$: Stringency index and continent are not independent (have an association).

We are performing a Chi-Square Test at the $\alpha$ = .05 level.

We have a chi-square distribution with our test statistic $\chi$^2 = 28610 and 835 degrees of freedom.

We have a P-value = 2.2e-16 > $\alpha$ = 0.05. Thus, we reject $H_0$. We have convincing evidence to suggest that stringency index and continent are not independent.


```{r Chi Square Plot}

#Why isn't visualization good

chisq_continent_stringency <- chisq.test(table(covid$continent, covid$stringency_index))

chisq_continent_stringency

ggchisqtest(chisq_continent_stringency, colaccept="lightsteelblue1", colreject="gray84", colstat="navyblue", alpha=0.05)


```



```{r Exploration of H3}

covid_continent_str_i <- covid %>%
  group_by(continent) %>%
  filter(!is.na(stringency_index)) %>%
  count(total_str_index = sum(stringency_index)) %>%
  mutate(avg_str_index = total_str_index/n) %>%
  arrange(desc(avg_str_index)) %>%
  select(continent, avg_str_index) %>%
  print(covid_continent_str_i)

```
In analyzing the average stringency indexes, there seem to be South America, North America, and Asia clumped at the highest stringency index, Africa being in the middle, and Oceania and Europe remaining way below them. This contradicts our hypothesis that Africa and South America would have the lowest stringency indexes and Oceania and Europe would be among the continents with the highest stringency index.




Based on the summary statistics, we maintain our hypothesis that stringency index and continents have a statistically significant relationship, however, we rehypothesize that the group of continents that are greater are South America, North America, Asia, and Africa, and the group of continents that are lower are Oceania and Europe.

$H_0$: The true mean stringency index of Europe, Asia, Oceania, and North America is equal to the true mean stringency index of Africa and South America

$H_1$: The true mean stringency index of Europe, Asia, Oceania, and North America is greater the true mean stringency index of Africa and South America



We are performing a CLT-based Hypothesis Test at the $\alpha$ = .05 level.


```{r Grouped Continents Test}

covid_si <- covid_final_date %>%
  filter(!is.na(stringency_index)) %>%
  mutate(grouped_continents = case_when(
    ((continent == "Oceania") | (continent == "Europe")) ~ "oc_eu",
    ((continent == "South America") | (continent == "Asia") ||
       (continent == "Africa") | (continent == "North America"))
    ~ "sa_na_as_af"))
 
#Group 1
covid_si %>% 
  filter(grouped_continents == "oc_eu") %>% 
  summarise(mean_si = mean(stringency_index)) %>% 
  print(covid_si)

#Group 2
covid_si %>% 
  filter(grouped_continents == "sa_na_as_af") %>% 
  summarise(mean_si = mean(stringency_index)) %>% 
  print(covid_si)

t.test(stringency_index ~ grouped_continents,
       data = covid_si,
       mu = 0,
       var.equal = FALSE,
       alternative = "less",
       conf.level = 0.95)


```







```{r Visualize Continent Stringency Indexes over Time}

covid_continent_str_i_2 <- covid %>%
  group_by(continent, date) %>%
  filter(!is.na(stringency_index)) %>%
  mutate(avg_str_index = mean(stringency_index))

#Make title more informative
ggplot(data = covid_continent_str_i_2, aes(x=date, y=avg_str_index, 
                                           color = continent)) +
  geom_line()+
  labs(x = "Date",
       y = "Mean Stringency Index",
       title = "Mean Stringency Index Over Time",
       subtitle = "Color by Continent")


```

While the past hypothesis tests looked at the final stringency index in our dataset, it may be interesting to see how stringency indexes change over time, and how that may affect cases. 

Our visualization highlights similar growth patterns in stringency index among the different continents with a significant rise in March, maximum in April, and slow decline afterward. One can see that Asia and Oceania were quicker to act and increase stringency index, having their stringency indexes around 20 by February (about double the other continents).

We've already observed from our correlation tables above that we don't have enough evidence to suggest there's a correlation between stringency index and the amount of covid cases. However, we now hypothesize that the reasone we failed to reject our first null hypothesis is that once cases exist, there is little you can do to contain the spread. Thus, we want to assess the possibility that stringency index is correlated to new cases per capita only in the early months of spread.

One downfall we see is that countries change their stringency index in reaction ot rising cases. This can give the illusion that they're more positively correlated than they may (not) be. Thus, we will create a new variable that tracks the factor by which new_cases is changing from day to day. We will check this correlation agains stringency index. After all, a high stringency index is designed to SLOW the spread of a virus - not stop it.

``` {r eda-stringency-early-cases}

# Make some sort of variable that tracks a factor by which new cases is growing each day.

n_days <- 90

covid_early <- covid %>%
  group_by(location) %>%
  filter(total_cases > 0 & !is.na(stringency_index) & !is.na(new_cases)) %>%
  slice(1:n_days) %>%
  select(continent, location, total_cases, new_cases, stringency_index)

covid_early

covid_early %>%
  summarize(cor_var_string = cor(stringency_index, new_cases))



```























### Manny

Hypothesis 2: We hypothesize that deaths per case will be largely determined by the GDP per capita, the number of citizens aged 65+, hospital beds per thousand, the human development index, and columns concerning pre-existing conditions (i.e. diabetes prevalence, cardiovascular death rate, etc).



```{r Visualize Deaths/Case Over Time}

#Could move this to exploratory data analysis


#Categories created based on 
# https://onlinelibrary.wiley.com/doi/full/10.1002/brb3.1755




covid <- covid %>% 
  mutate(hdi_columns = case_when(
    human_development_index >= 0.8 ~ "Very High",
    human_development_index < 0.8 & human_development_index >= 0.7 ~ "High",
    human_development_index < 0.7 & human_development_index >= 0.55 ~ "Medium",
    human_development_index < 0.55~ "Low"
  )) %>% 
  mutate(deaths_per_case = total_deaths/total_cases)

  
covid_deaths_case <- covid %>%
  group_by(date) %>%
  filter(!is.na(deaths_per_case)) %>%
  mutate(avg_deaths_case = mean(deaths_per_case))

#Make title more informative
ggplot(data = covid_deaths_case, aes(x=date, y=avg_deaths_case)) +
  geom_line()+
  labs(x = "Date",
       y = "Deaths/Case",
       title = "Deaths Per Case Over Time")


covid_deaths_hdi <- covid %>%
  group_by(hdi_columns, date) %>%
  filter(!is.na(deaths_per_case) & !is.na(human_development_index)) %>%
  mutate(avg_deaths_case = mean(deaths_per_case))



#Make title more informative
ggplot(data = covid_deaths_hdi, aes(x=date, y=avg_deaths_case, 
                                    color = hdi_columns)) +
  geom_line()+
  labs(x = "Date",
       y = "Deaths/Case",
       title = "Mean Deaths Per Case Over Time",
       subtitle = "Color by Human Development Index Category")



```


``` {r manny}

# In order to accurately compare between countries, mutate deaths per case
covid_final_date <- covid_final_date %>% 
  mutate(deaths_per_case = total_deaths/total_cases)

# Get rid of null values for numeric variables
filtered_covid <- covid_final_date %>%
  filter(gdp_per_capita > 0) %>% 
  filter(aged_65_older > 0) %>% 
  filter(hospital_beds_per_thousand > 0) %>% 
  filter(human_development_index >0)
  
ggplot(data = filtered_covid, mapping = aes(x = gdp_per_capita, 
                                             y = deaths_per_case)) +
  geom_point()
r_gdp <- glance(lm(deaths_per_case ~ gdp_per_capita, 
                   data = filtered_covid))$r.squared



ggplot(data = filtered_covid, mapping = aes(x = aged_65_older, 
                                             y = deaths_per_case)) +
  geom_point()
r_seniors <- glance(lm(deaths_per_case ~ aged_65_older, 
                       data = filtered_covid))$r.squared



ggplot(data = filtered_covid, mapping = aes(x = hospital_beds_per_thousand, 
                                             y = deaths_per_case)) +
  geom_point()
r_hospital <- glance(lm(deaths_per_case ~ hospital_beds_per_thousand, 
            data = filtered_covid))$r.squared



ggplot(data = filtered_covid, mapping = aes(x = human_development_index, 
                                             y = deaths_per_case)) +
  geom_point()
r_dev <- glance(lm(deaths_per_case ~ human_development_index, 
            data = filtered_covid))$r.squared

r_gdp
r_seniors
r_hospital
r_dev


```


Given the visualizations between all the continuous numerical variables and deaths per case, as well as the correlations from linear regressions, each of these variables by themselves seems to have little to no impact at all on the deaths per case from COVID. However it very well may be that that the combination between these variables as well as the possibility for interaction terms between them could predict COVID much better.

Hypothesis 2: We hypothesize that deaths per case will be largely determined by the GDP per capita, the number of citizens aged 65+, hospital beds per thousand, the human development index, and columns concerning pre-existing conditions (i.e. diabetes prevalence, cardiovascular death rate, etc).


### Frankie
``` {r frankie}

covid_final_date_deaths <- covid_final_date %>% 
  filter(!is.na(total_deaths) & !is.na(total_cases)) %>% 
  mutate(deaths_per_case = total_deaths / total_cases)

covid_num <- select_if(covid_final_date_deaths, is.numeric) # Gets all numeric variables
covid_cat <- select_if(covid_final_date_deaths, is.character) # Gets all categorical variables
covid_cat$deaths_per_case = covid_final_date_deaths$deaths_per_case

col_names <- colnames(covid_num)

cors_deaths <- c(rep(NA, ncol(covid_num)))
for(i in 1:ncol(covid_num)) {
  isolated <- covid_num %>%
    filter(!is.na(deaths_per_case) & !is.na(covid_num[[i]])) %>%
    select(col_names[i], deaths_per_case)

  cors_deaths[i] <- cor(isolated)[2]
}

correlations_deaths <- tibble(cors_deaths)
correlations_deaths$other_var = col_names
correlations_deaths <- correlations_deaths %>%
  filter(!str_detect(other_var,'cases') & !str_detect(other_var,'deaths')) %>% 
  arrange(desc(abs(cors_deaths))) %>% 
  print(correlations_deaths)

#str_detect from
#https://stackoverflow.com/questions/22850026/filter-rows-which-contain-a-certain-string

# Hypothesis test (H_1)
# cor.test(covid$stringency_index, covid$total_cases_per_cap, alternative="less")




```

```{r Determine Variables with Statisticall Significant Relationships with Deaths/Case}


#Trying to run loop to determine variables with statistically significant correlations

# for(i in 1:ncol(covid_num)) {
#   isolated <- covid_final_date_deaths %>%
#     filter(!is.na(deaths_per_case) & !is.na(col_names[i]))
#   cor.test(isolated$as.matrix(col_names[i]), isolated$total_cases_per_cap)
# }



```




Variables with st

```{r Mulitiple Linear Regression}

#Could try running a bunch of lines and then seeing which has maximum adjust R^2

m_main <- lm(deaths_per_case ~ male_smokers + diabetes_prevalence + hospital_beds_per_thousand + aged_70_older + population_density + aged_65_older + population + female_smokers + extreme_poverty + stringency_index + human_development_index + gdp_per_capita + life_expectancy + cardiovasc_death_rate + handwashing_facilities + median_age, data = covid_final_date_deaths)

m_main %>%
 tidy()

glance(m_main) %>% 
  select(r.squared, adj.r.squared)


```



In taking the most variables with the largest correlations with the greatest correlations, we can create a linear regression model to attempt to predict a country's deaths per cases based on its data and demographics. 



```{r Load in GLM Net}

install.packages("glmnet", repos = "http://cran.us.r-project.org")


```

```{r Add Libraries}

library(glmnet)
library(caret)

```


```{r ML Test Summarize Data}

# Running penalized regression (Lasso) based on guidance of:
# https://www.pluralsight.com/guides/linear-lasso-and-ridge-regression-with-r

set.seed(100) 

index = sample(1:nrow(covid_final_date_deaths), 0.7*nrow(covid_final_date_deaths)) 

train = covid_final_date_deaths[index,]
test = covid_final_date_deaths[-index,]


cols = c('male_smokers', 'diabetes_prevalence', 'hospital_beds_per_thousand',  'aged_70_older', 'population_density','aged_65_older', 'population','female_smokers', 'extreme_poverty','stringency_index', 'human_development_index', 'gdp_per_capita', 'life_expectancy', 'cardiovasc_death_rate', 'handwashing_facilities', 'median_age')

pre_proc_val <- preProcess(train[,cols], method = c("center", "scale"))

train[,cols] = predict(pre_proc_val, train[,cols])
test[,cols] = predict(pre_proc_val, test[,cols])

summary(train)



```


```{r ML Split Data}

cols_reg = c('male_smokers', 'diabetes_prevalence', 'hospital_beds_per_thousand',  'aged_70_older', 'population_density','aged_65_older', 'population','female_smokers', 'extreme_poverty','stringency_index', 'human_development_index', 'gdp_per_capita', 'life_expectancy', 'cardiovasc_death_rate', 'handwashing_facilities', 'median_age', 'deaths_per_case')


dummies <- dummyVars(deaths_per_case ~ ., data = covid_final_date_deaths[,cols_reg])
train_dummies = predict(dummies, newdata = train[,cols_reg])
test_dummies = predict(dummies, newdata = test[,cols_reg])


#Need to fix- y and x diff lengths

x = as.matrix(train_dummies)
y_train = covid_final_date_deaths$deaths_per_case

x_test = as.matrix(test_dummies)
y_test = covid_final_date_deaths$deaths_per_case




```

```{r ML Find Lambda}

lambdas <- 10^seq(2, -3, by = -.1)

lasso_reg <- cv.glmnet(x, y_train, alpha = 1, lambda = lambdas, standardize = TRUE, nfolds = 5)

lambda_best <- lasso_reg$lambda.min 
lambda_best


```

```{r ML Run Lasso}


lasso_model <- glmnet(x, y_train, alpha = 1, lambda = lambda_best, standardize = TRUE)


```

```{r Eval Lasso}

eval_results <- function(true, predicted, df) {
  SSE <- sum((predicted - true)^2)
  SST <- sum((true - mean(true))^2)
  R_square <- 1 - SSE / SST
  RMSE = sqrt(SSE/nrow(df))
}

predictions_train <- predict(lasso_model, s = lambda_best, newx = x)
eval_results(y_train, predictions_train, train)

predictions_test <- predict(lasso_model, s = lambda_best, newx = x_test)
eval_results(y_test, predictions_test, test)


```





```{r Hypothesis 1 Visualization}

#Hypothesis for cases per capita strong negative correlation to stringency index.

#Make variable depending on states that had highest final cases per death

# covid_cases_stringency <- covid %>%
#   group_by(continent, date) %>%
#   filter(!is.na(stringency_index)) %>%
#   mutate(avg_str_index = mean(stringency_index))
# 
# #Make title more informative
# ggplot(data = covid_continent_str_i_2, aes(x=date, y=avg_str_index,
#                                            color = continent)) +
#   geom_line()+
#   labs(x = "Date",
#        y = "Mean Stringency Index",
#        title = "Mean Stringency Index Over Time",
#        subtitle = "Color by Continent")

ggplot(data = covid_continent_str_i_2, aes(x=date, y=total_cases_per_cap,
                                           color = continent)) +
  geom_line()+
  labs(x = "Date",
       y = "Mean Stringency Index",
       title = "Mean Stringency Index Over Time",
       subtitle = "Color by Continent")



```




###Conclusion













